Random Generation of Rich \\ Abstract Data Type Values

### 2019-01-30 paper/synthesis.tex
Added syntax highlighting (config is in notation.fmt)

Vector $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given level, provided that we
generated a constructor of kind $i$ at the previous level.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions \cite{DBLP:conf/haskell/MistaRH18}---specially the recursive
fields---as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{Join, Text} = 2\cdot p_{Text} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{Text}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

@@

It is our realization that it suffices to simply consider \emph{each different
  pattern matching and function call as a kind of individual on its own}.

@@

In that manner, we can extend our mean matrix $M$ by adding a row and a column
for each different pattern matching and function call as follows:

@@

@@

\begin{equation*}
  \hspace{20pt}
  M = \ \

@@

    [ baseline=-0.65ex

@@

    \fill[red, opacity=0.075]

@@

    -|| ([xshift=-1pt]m-1-3.north east)
    -|| ([xshift=-1pt, yshift=1pt]m-3-3.south east)
    -|| ([xshift=1pt, yshift=1pt]m-3-1.south west)

@@

    \fill[blue, opacity=0.075]

@@

    -|| ([xshift=1pt]m-1-3.north east)
    -|| ([xshift=0pt, yshift=-1pt]m-1-9.south east)
    -|| ([xshift=0pt, yshift=0pt]m-9-9.south east)

@@

    \node[above,text depth=1pt] at (m-1-1.north) {$\scriptstyle C_1$};
    \node[above,text depth=1pt] at (m-1-2.north) {$\scriptstyle \cdots$};
    \node[above,text depth=1pt] at (m-1-3.north) {$\scriptstyle C_i$};
    \node[above,text depth=1pt] at (m-1-4.north) {$\scriptstyle P_1$};
    \node[above,text depth=1pt] at (m-1-5.north) {$\scriptstyle \cdots$};
    \node[above,text depth=1pt] at (m-1-6.north) {$\scriptstyle P_j$};
    \node[above,text depth=1pt] at (m-1-7.north) {$\scriptstyle F_1$};
    \node[above,text depth=1pt] at (m-1-8.north) {$\scriptstyle \cdots$};
    \node[above,text depth=1pt] at (m-1-9.north) {$\scriptstyle F_k$};

@@

    \node[left,overlay] at (m-1-1.west) {$\scriptstyle C_1$};
    \node[left,overlay] at ([xshift=-4pt, yshift=2pt]m-2-1.west) {$\scriptstyle \vdots$};
    \node[left,overlay] at (m-3-1.west) {$\scriptstyle C_i$};
    \node[left,overlay] at (m-4-1.west) {$\scriptstyle P_1$};
    \node[left,overlay] at ([xshift=-4pt, yshift=2pt]m-5-1.west) {$\scriptstyle \vdots$};
    \node[left,overlay] at (m-6-1.west) {$\scriptstyle P_j$};
    \node[left,overlay] at (m-7-1.west) {$\scriptstyle F_1$};
    \node[left,overlay] at ([xshift=-4pt, yshift=2pt]m-8-1.west) {$\scriptstyle \vdots$};
    \node[left,overlay] at (m-9-1.west) {$\scriptstyle F_k$};

@@

    \draw[densely dashed] ([xshift=-2pt]m-3-1.south west) -- ([xshift=2pt]m-3-9.south east);
    \draw[densely dashed] ([xshift=-2pt]m-6-1.south west) -- ([xshift=2pt]m-6-9.south east);
    \draw[densely dashed] ([yshift=2pt]m-1-3.north east) -- ([yshift=-2pt]m-9-3.south east);
    \draw[densely dashed] ([yshift=2pt]m-1-6.north east) -- ([yshift=-2pt]m-9-6.south east);

@@

    \draw[dotted, semithick] (m-6-1.west) -- ([xshift=4pt]m-6-7.west);
    \draw[dotted, semithick] (m-1-7.north) -- ([yshift=-3pt]m-6-7.north);

@@

\end{equation*}

Symbol $C_1 \cdots C_i$ denotes constructors, $P_1 \cdots P_j$ pattern
matchings, and $F_1 \cdots F_k$ function calls.
%
The red-light colored matrix is what we had before where the light-blue colored
cells are new---we encourage readers to obtain a colored copy of this work.
%
% To extend $M$ to consider the new possible constructions, we need to calculate
% $M_{i,j}$ for every new combination of ``father'' and ``son'' constructions $i$
% and $j$, respectively.
% %

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of recursive sub-terms of each construction that we
generate (i.e., data constructors, pattern matching values or function calls),
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}
%
The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as follows (where dashed lines indicate the composition of the
diffent sub-vectors that are calculated separately):

\begin{equation*}
  \hspace{20pt}
  \beta = \qquad \qquad \
  \begin{tikzpicture}
    [ baseline=-0.65ex
    , every left delimiter/.style={xshift=.5em}
    , every right delimiter/.style={xshift=-.5em}
    ]
    \matrix
    [ matrix of math nodes
    , column sep=0ex
    , row sep=-1pt
    , ampersand replacement=\&
    , left delimiter={[}
    , right delimiter={]}
    ] (v)
    {
      0 \\ % Text
      0 \\ % Sing
      1 \\ % Tag
      2 \\ % Join
      0 \\ % simplify#1
      2 \\ % simplify#1
      0 \\ % hr
      1 \\ % div
      1 \\ % bold
    };
    \draw[densely dashed] ([xshift=-3pt]v-4-1.south west) -- ([xshift=3pt]v-4-1.south east);
    \draw[densely dashed] ([xshift=-3pt]v-6-1.south west) -- ([xshift=3pt]v-6-1.south east);
    \node[left,overlay] at ([xshift=-2pt]v-1-1.west) {$\scriptstyle Text$};
    \node[left,overlay] at ([xshift=-2pt]v-2-1.west) {$\scriptstyle Single$};
    \node[left,overlay] at ([xshift=-2pt]v-3-1.west) {$\scriptstyle Tag$};
    \node[left,overlay] at ([xshift=-2pt]v-4-1.west) {$\scriptstyle Join$};
    \node[left,overlay] at ([xshift=-2pt]v-5-1.west) {$\scriptstyle simplify\#1$};
    \node[left,overlay] at ([xshift=-2pt]v-6-1.west) {$\scriptstyle simplify\#2$};
    \node[left,overlay] at ([xshift=-2pt]v-7-1.west) {$\scriptstyle hr$};
    \node[left,overlay] at ([xshift=-2pt]v-8-1.west) {$\scriptstyle div$};
    \node[left,overlay] at ([xshift=-2pt]v-9-1.west) {$\scriptstyle bold$};
  \end{tikzpicture}
  \hspace{20pt}
  \mathcal{P} =
  \begin{tikzpicture}
    [ baseline=-0.65ex
    , every left delimiter/.style={xshift=.5em}
    , every right delimiter/.style={xshift=-.5em}
    ]
    \matrix
    [ matrix of math nodes
    , column sep=0ex
    , row sep=-1pt
    , ampersand replacement=\&
    , left delimiter={[}
    , right delimiter={]}
    ] (v)
    {
      p_{Text} \\
      p_{Single} \\
      p_{Tag}  \\
      p_{Join} \\
      p_{simplify\#1} \\
      p_{simplify\#2} \\
      p_{hr} \\
      p_{div} \\
      p_{bold} \\
    };
    \draw[densely dashed] ([xshift=-15pt]v-4-1.south west) -- ([xshift=15pt]v-4-1.south east);
    \draw[densely dashed] ([xshift=-3pt]v-6-1.south west) -- ([xshift=3pt]v-6-1.south east);
  \end{tikzpicture}
\end{equation*}

>>>>

Vector $E[G_0]$ represents the initial distribuition of constructors that our
generator produces, which simply consists of the generation probability of each
one.
%
The interesting aspect of the prediction mechanism is encoded in the matrix $M$,
known as a the \emph{mean matrix} of this stochastic process.
%
$M$ is a squared matrix with as rows and columns as different data constructors
involved in the generation process.
%
Each element $M_{i,j}$ of this matrix encodes the average number of data
constructors of kind $j$ that gets generated in a given level, provided that we
generated a constructor of kind $i$ at the previous level.
%
In this sense, this matrix encodes the ``branching'' behavior of our random
generation from one generation to the next one.
%
Each element of the matrix can be automatically calculated by exploiting ADT
definitions \cite{DBLP:conf/haskell/MistaRH18}
%
% ---specially the recursive fields---
%
as well as the individual probability of generating each constructor.
%
For instance, the average number of |Text| data constructors that we will
generate provided that we generated a |Join| data constructor on the previous
level is calculated as follows:
%
$$ M_{Join, Text} = 2\cdot p_{Text} $$
%
where $2$ is the number of holes present when generating a partial ADT value
|Join| (i.e., |Join hole hole|) and $p_{Text}$ is the probability of
individually generating the constructor |Text|.
%
This pattern can be used to build the rest of the mean matrix as well.

@@

Our realization is that it suffices to consider \emph{each different pattern
  matching and function call as a kind of individual on its own}.

@@

In that manner, we can extend our mean matrix $M$ adding a row and a column for
each different pattern matching and function call as shown in Fig.
\ref{fig:meanmatrix}.

@@

\begin{wrapfigure}{r}{0.26\textwidth}
  \vspace{-8pt}

@@

  \centering
  \resizebox{4.5cm}{!}{

@@

      [ baseline=-0.5ex
      % , transform canvas={scale=0.9}

@@

      \fill[red, opacity=0.15]

@@

      -|| (m-1-3.north east)
      -|| (m-3-3.south east)
      -|| ([xshift=1pt]m-3-1.south west)

@@

      \fill[blue, opacity=0.15]

@@

      -|| (m-1-3.north east)
      -|| ([yshift=-1pt]m-1-9.south east)
      -|| (m-9-9.south east)

@@

      \node[above,text depth=1pt] at (m-1-1.north) {$ C_1$};
      \node[above,text depth=1pt] at (m-1-2.north) {$ \cdots$};
      \node[above,text depth=1pt] at (m-1-3.north) {$ C_i$};
      \node[above,text depth=1pt] at (m-1-4.north) {$ P_1$};
      \node[above,text depth=1pt] at (m-1-5.north) {$ \cdots$};
      \node[above,text depth=1pt] at (m-1-6.north) {$ P_j$};
      \node[above,text depth=1pt] at (m-1-7.north) {$ F_1$};
      \node[above,text depth=1pt] at (m-1-8.north) {$ \cdots$};
      \node[above,text depth=1pt] at (m-1-9.north) {$ F_k$};

@@

      \node[left,overlay] at (m-1-1.west) {$ C_1$};
      \node[left,overlay] at ([xshift=-4pt, yshift=2pt]m-2-1.west) {$ \vdots$};
      \node[left,overlay] at (m-3-1.west) {$ C_i$};
      \node[left,overlay] at (m-4-1.west) {$ P_1$};
      \node[left,overlay] at ([xshift=-4pt, yshift=2pt]m-5-1.west) {$ \vdots$};
      \node[left,overlay] at (m-6-1.west) {$ P_j$};
      \node[left,overlay] at (m-7-1.west) {$ F_1$};
      \node[left,overlay] at ([xshift=-4pt, yshift=2pt]m-8-1.west) {$ \vdots$};
      \node[left,overlay] at (m-9-1.west) {$ F_k$};

@@

      \draw[densely dashed, semithick] ([xshift=-2pt]m-3-1.south west) -- ([xshift=2pt]m-3-9.south east);
      \draw[densely dashed, semithick] ([xshift=-2pt]m-6-1.south west) -- ([xshift=2pt]m-6-9.south east);
      \draw[densely dashed, semithick] ([yshift=2pt]m-1-3.north east) -- ([yshift=-2pt]m-9-3.south east);
      \draw[densely dashed, semithick] ([yshift=2pt]m-1-6.north east) -- ([yshift=-2pt]m-9-6.south east);

@@

      \draw[dotted, thick] (m-6-1.west) -- ([xshift=4pt]m-6-7.west);
      \draw[dotted, thick] (m-1-7.north) -- ([yshift=-3pt]m-6-7.north);

@@

}
  \caption{Mean matrix $M$ including pattern matching and function calls
    information.}
  \label{fig:meanmatrix}
  \vspace{-6pt}
\end{wrapfigure}
Symbol $C_1\! \cdots C_i$ denotes constructors, $P_1\! \cdots P_j$ pattern
matchings, and $F_1\! \cdots F_k$ function calls.
The red-light colored matrix is what we had before where the light-blue colored
cells are new---we encourage readers to obtain a colored copy of this work.
% To extend $M$ to consider the new possible constructions, we need to calculate
% $M_{i,j}$ for every new combination of ``father'' and ``son'' constructions $i$
% and $j$, respectively.
% %

@@

As another contribution, we found that the whole prediction process can be
factored in terms of two vectors $\beta$ and $\mathcal{P}$, such that $\beta$
represents the number of recursive sub-terms of each construction that we
generate (i.e., data constructors, pattern matching values or function calls),
whereas $\mathcal{P}$ simply represents the probability of generating that
partial ADT value.% (this probability can be easily calculated based on the
% generation frequencies used in the random generator).
%
Then, the equation \ref{eqn:prediction} can be rewriten as:
%
\begin{align*}
  E[G_n]^T = \beta^T \cdot (\beta \cdot \mathcal{P})^{n}
\end{align*}
%
The vector $\beta$ can be obtained by analyzing the shape (namely holes) of each
source of structural information that we consider, and joining them together.
%
For instance, $\beta$ and $\mathcal{P}$ for our generation specification of HTML
values are as shown in Fig. \ref{fig:vectors} (where dashed lines indicate the
composition of the diffent sub-vectors that are calculated separately).

\begin{figure}[t]
  \resizebox{!}{2cm}{
\begin{equation*}
  \hspace{20pt}
  \beta = \qquad \qquad \
  \begin{tikzpicture}
    [ baseline=-0.65ex
    , every left delimiter/.style={xshift=.5em}
    , every right delimiter/.style={xshift=-.5em}
    ]
    \matrix
    [ matrix of math nodes
    , column sep=0ex
    , row sep=-1pt
    , ampersand replacement=\&
    , left delimiter={[}
    , right delimiter={]}
    ] (v)
    {
      0 \\ % Text
      0 \\ % Sing
      1 \\ % Tag
      2 \\ % Join
      0 \\ % simplify#1
      2 \\ % simplify#1
      0 \\ % hr
      1 \\ % div
      1 \\ % bold
    };
    \draw[densely dashed] ([xshift=-3pt]v-4-1.south west) -- ([xshift=3pt]v-4-1.south east);
    \draw[densely dashed] ([xshift=-3pt]v-6-1.south west) -- ([xshift=3pt]v-6-1.south east);
    \node[left,overlay] at ([xshift=-2pt]v-1-1.west) {$\scriptstyle Text$};
    \node[left,overlay] at ([xshift=-2pt]v-2-1.west) {$\scriptstyle Single$};
    \node[left,overlay] at ([xshift=-2pt]v-3-1.west) {$\scriptstyle Tag$};
    \node[left,overlay] at ([xshift=-2pt]v-4-1.west) {$\scriptstyle Join$};
    \node[left,overlay] at ([xshift=-2pt]v-5-1.west) {$\scriptstyle simplify\#1$};
    \node[left,overlay] at ([xshift=-2pt]v-6-1.west) {$\scriptstyle simplify\#2$};
    \node[left,overlay] at ([xshift=-2pt]v-7-1.west) {$\scriptstyle hr$};
    \node[left,overlay] at ([xshift=-2pt]v-8-1.west) {$\scriptstyle div$};
    \node[left,overlay] at ([xshift=-2pt]v-9-1.west) {$\scriptstyle bold$};
  \end{tikzpicture}
  \hspace{30pt}
  \mathcal{P} =
  \begin{tikzpicture}
    [ baseline=-0.65ex
    , every left delimiter/.style={xshift=.5em}
    , every right delimiter/.style={xshift=-.5em}
    ]
    \matrix
    [ matrix of math nodes
    , column sep=0ex
    , row sep=-1pt
    , ampersand replacement=\&
    , left delimiter={[}
    , right delimiter={]}
    ] (v)
    {
      p_{Text} \\
      p_{Single} \\
      p_{Tag}  \\
      p_{Join} \\
      p_{simplify\#1} \\
      p_{simplify\#2} \\
      p_{hr} \\
      p_{div} \\
      p_{bold} \\
    };
    \draw[densely dashed] ([xshift=-15pt]v-4-1.south west) -- ([xshift=15pt]v-4-1.south east);
    \draw[densely dashed] ([xshift=-3pt]v-6-1.south west) -- ([xshift=3pt]v-6-1.south east);
  \end{tikzpicture}
\end{equation*}
}
\vspace{-5pt}
\caption{Prediction vectors of our |Html| generation specification.}
\label{fig:vectors}
\vspace{-10pt}
\end{figure}
